---
title: "Lab3"
output: html_document
date: "2024-10-04"
---
Variable Distributions (2%)
Using the EPI results dataset to perform the following:
1. Derive 2 subsets each for a particular region
  1.1. Plot histograms for a variable of your choice for both regions with density             lines overlayed
  
  1.2. Plot QQ plots for both variables compared to known probability distributions.
```{r}
setwd("~/RPI/Coterm_Fall_2024/Data_Analytics/Michael-Dong-Data-Analytics/Lab3/")
EPI_data <- read.csv("epi2024results_DA_F24_lab03.csv")
# View(EPI_data)
set.seed(123)
attach(EPI_data)

Asia_Pacific <- EPI_data[EPI_data$region == "Asia-Pacific",]
Sub_Sahara <- EPI_data[EPI_data$region == "Sub-Saharan Africa",]

# summary(Asia_Pacific)
# summary(Sub_Sahara)

hist(Asia_Pacific$EPI, seq(20., 70., 1.0), prob=TRUE)
lines(density(Asia_Pacific$EPI,na.rm=TRUE,bw=1.))

hist(Sub_Sahara$EPI, seq(20., 60., 1.0), prob=TRUE)
lines(density(Sub_Sahara$EPI,na.rm=TRUE,bw=1.))


qqplot(rnorm(250), Asia_Pacific$EPI, xlab = "Q-Q plot for norm dsn")
qqline(Asia_Pacific$EPI)

qqplot(rt(250, df = 5) ,Sub_Sahara$EPI, xlab = "Q-Q plot for norm dsn")
qqline(Sub_Sahara$EPI)
```


Linear Models (3%)
Using the EPI results dataset to perform the following:
2. Fit linear models as follows:
  2.1. Choose a subset of 5 variables (excluding EPI) and using the formula
        EPI~VAR1+VAR2+VAR3+VAR4+VAR5, fit a linear model and identify which variable          most significantly influences EPI. Plot that variable with another and overlay        the fitted line.
        
  2.2. Repeat the previous model with a subset of 1 region and in 1-2 sentences              explain which model is a better fit and why you think that is the case.
  
  2.2 Answer: The model that includes all regions because there are more data points                to train from which allows better prediction models.
```{r}
library(ggplot2)
setwd("~/RPI/Coterm_Fall_2024/Data_Analytics/Michael-Dong-Data-Analytics/Lab3/")
EPI_data <- read.csv("epi2024results_DA_F24_lab03.csv")
# View(EPI_data)
attach(EPI_data)
set.seed(123)

Asia_Pacific <- na.omit(EPI_data[EPI_data$region == "Asia-Pacific",])
# Sub_Sahara <- na.omit(EPI_data[EPI_data$region == "Sub-Saharan Africa",])
# East_Euro <- na.omit(EPI_data[EPI_data$region == "Eastern Europe",])
# Global_West <- na.omit(EPI_data[EPI_data$region == "Global West",])

# lin.mod.epi_asia <- lm(EPI~BDH+ECO+VOE+WWR+MKP, Asia_Pacific)
# lin.mod.epi_asia <- lm(EPI~BDH+ECO+VOE+WWR+MKP, EPI_data)

# summary(lin.mod.epi_asia)

ggplot(EPI_data, aes(x = ECO, y = BDH)) +
  geom_point() +
  stat_smooth(method = "lm")

ggplot(Asia_Pacific, aes(x = ECO, y = BDH)) +
  geom_point() +
  stat_smooth(method = "lm")
```


Classification (kNN) (3%)
Using the EPI results dataset to perform the following:
3. Train 2 kNN models using ”region” as the class label as follows:
  3.1. Choose a subset of 5 variables and filter the subset by region keeping 3             regions out of 8 (representing 3 classes), then train a kNN model to predict the       region based on these variables. Evaluate the model using a contingency matrix        and calculate the accuracy of correct classifications.
  
  3.2. Repeat the previous model with the same variables for another set of 3 other          regions and evaluate. In 1- 2 sentences explain which model is better and why         you think that is the case.
  
  3.2 Answer: The first model (sub_region 1) because it has a higher accuracy. This is
              the case because the subset of regions in the second model may not be   
              closely correlated with each other due to location or current state of 
              situtations. 
  
```{r}
setwd("~/RPI/Coterm_Fall_2024/Data_Analytics/Michael-Dong-Data-Analytics/Lab3/")

library(class)
set.seed(123)
EPI_data <- read.csv("epi2024results_DA_F24_lab03.csv")
filtered_EPI_data <- EPI_data[,c(1:8,14,15)]
sub_region1 <- filtered_EPI_data[filtered_EPI_data$region %in% c("Asia-Pacific", "Eastern Europe", "Global West"),]

s_region <- sample(66,46)

#  ## create train & test sets based on sampled indexes
sub_region1.train <- sub_region1[s_region,]
sub_region1.test <- sub_region1[-s_region,]

sub_region1.train <- sub_region1.train[complete.cases(sub_region1.train),]
sub_region1.test <- sub_region1.test[complete.cases(sub_region1.test),]

# train model & predict
KNNpred <- knn(train = sub_region1.train[6:10], test = sub_region1.test[6:10], cl = sub_region1.train$region, k = 5)
# create contingency table/ confusion matrix
contingency.table <- table(KNNpred,sub_region1.test$region)

contingency.table

contingency.matrix = as.matrix(contingency.table)
sum(diag(contingency.matrix))/length(sub_region1.test$region)
accuracy <- c()
ks <- c(3,5,7,9,11)
for (k in ks) {
KNNpred <- knn(train = sub_region1.train[6:10], test = sub_region1.test[6:10], cl = sub_region1.train$region, k = k)
cm = as.matrix(table(Actual=KNNpred, Predicted = sub_region1.test$region, dnn=list('predicted','actual')))

accuracy <- c(accuracy,sum(diag(cm))/length(sub_region1.test$region))
}
plot(ks,accuracy,type = "b", main = "sub_region 1")

#############Second subset of regions
sub_region2 <- filtered_EPI_data[filtered_EPI_data$region %in% c("Sub-Saharan Africa", "Latin America & Caribbean", "Greater Middle East"),]
s_region2 <- sample(94,65)


#  ## create train & test sets based on sampled indexes
sub_region2.train <- sub_region2[s_region2,]
sub_region2.test <- sub_region2[-s_region2,]

sub_region2.train <- sub_region2.train[complete.cases(sub_region2.train),]
sub_region2.test <- sub_region2.test[complete.cases(sub_region2.test),]

# train model & predict
KNNpred <- knn(train = sub_region2.train[6:10], test = sub_region2.test[6:10], cl = sub_region2.train$region, k = 5)
# create contingency table/ confusion matrix
contingency.table <- table(KNNpred,sub_region2.test$region)

contingency.table

contingency.matrix = as.matrix(contingency.table)
sum(diag(contingency.matrix))/length(sub_region2.test$region)
accuracy <- c()
ks <- c(3,5,7,9,11)
for (k in ks) {
KNNpred <- knn(train = sub_region2.train[6:10], test = sub_region2.test[6:10], cl = sub_region2.train$region, k = k)
cm = as.matrix(table(Actual=KNNpred, Predicted = sub_region2.test$region, dnn=list('predicted','actual')))

accuracy <- c(accuracy,sum(diag(cm))/length(sub_region2.test$region))
}
plot(ks,accuracy,type = "b", main = "sub_region 2")
```


Clustering (3%)
Using the EPI results dataset to perform the following:
1. Fit a k-means model for a subset of 5 variables for 2 different groups of regions        (3 each)
  1.1. Compare the performance of the models using their within cluster sum of                squares.
  
  1.2. In a loop, fit kmeans models for both subsets using multiple values of k. Plot          WCSS across k values. In 1-2 sentences explain which model is better and why            you think that is the case.
  
  1.2 Answer: Model 1 (group 1) because it has a lower WSS range meaning points are                 grouped much tighter together as number of clusters increase vs Model 2.               This might be the case because the subset of model 2 might have a lot of               wide-ranging points resulting in less consistent points around a                      cluster's centroid 
```{r}
#EPI, ECO, BDH, TBN, TKP
setwd("~/RPI/Coterm_Fall_2024/Data_Analytics/Michael-Dong-Data-Analytics/Lab3/")
library(rpart)
library(ggplot2)
library(rpart.plot)

EPI_data <- read.csv("epi2024results_DA_F24_lab03.csv")
filtered_EPI_data <- EPI_data[,c(1:8,14,15)]

group1 <- filtered_EPI_data[filtered_EPI_data$region %in% c("Asia-Pacific", "Eastern Europe", "Global West"),]

group2 <- filtered_EPI_data[filtered_EPI_data$region %in% c("Sub-Saharan Africa", "Latin America & Caribbean", "Greater Middle East"),]

# set seed for random number generator
set.seed(123)
# run k-means
group1.km <- kmeans(group1[,6:10], centers = 3)
wss <- c()
wss <- c(wss,group1.km$tot.withinss)
wss

group2.km <- kmeans(group2[,6:10], centers = 3)
wss2 <- c()
wss2 <- c(wss2,group2.km$tot.withinss)
wss2

wss <- c()
ks <- c(3,5,7,9)
for (k in ks) {
group1.km <- kmeans(group1[,6:10], centers = k)
wss <- c(wss,group1.km$tot.withinss)
}
plot(ks,wss,type = "b", main ="group 1")

wss2 <- c()
ks <- c(3,5,7,9)
for (k in ks) {
group2.km <- kmeans(group2[,6:10], centers = k)
wss2 <- c(wss2,group2.km$tot.withinss)
}
plot(ks,wss2,type = "b", main = "group 2")
```

